---
title: "Troubleshooting SpectroTune"
author: "John Albechaalany"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting SpectroTune}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

# Troubleshooting SpectroTune

This vignette addresses common issues and provides solutions for problems you might encounter when using SpectroTune.

```{r load}
library(spectrotune)
library(ggplot2)
```

## Installation Issues

### Package Installation Problems

**Problem**: `Error in loadNamespace(x) : there is no package called 'devtools'`

**Solution**:
```{r install_devtools, eval = FALSE}
# Install required packages
install.packages(c("devtools", "roxygen2"))

# If on Windows, install Rtools
# Download from: https://cran.rstudio.com/bin/windows/Rtools/
```

**Problem**: `Warning: cannot remove prior installation of package 'roxygen2'`

**Solution**:
```{r fix_roxygen, eval = FALSE}
# 1. Restart R/RStudio
# 2. Remove package manually
remove.packages("roxygen2")

# 3. Delete lock folders (Windows)
# Navigate to: C:\Users\[username]\AppData\Local\R\win-library\[version]\
# Delete: 00LOCK\roxygen2\ and roxygen2\ folders

# 4. Reinstall
install.packages("roxygen2", type = "binary")
```

### Permission Denied Errors

**Problem**: `Permission denied` when installing packages

**Solution**:
```{r fix_permissions, eval = FALSE}
# Check library paths
.libPaths()

# Install to user library
install.packages("package_name", lib = .libPaths()[1])

# Or run R as administrator (Windows)
```

## Data Format Issues

### Column Name Problems

**Problem**: `Error: TARGET_COL %in% names(raw_df) is not TRUE`

**Solution**:
```{r fix_column_names, eval = FALSE}
# Check column names
names(your_data)

# Ensure target column exists and is properly named
# The function will automatically detect and rename target columns
# Make sure your target column name is consistent

# Example: If your target is "Age" but column is "age"
names(your_data)[names(your_data) == "age"] <- "Age"
```

### Spectral Column Format

**Problem**: Spectral columns not recognized

**Solution**:
```{r fix_spectral_columns, eval = FALSE}
# Check if spectral columns are numeric
str(your_data)

# Convert to numeric if needed
spectral_cols <- grep("^[0-9]+$", names(your_data), value = TRUE)
your_data[spectral_cols] <- lapply(your_data[spectral_cols], as.numeric)

# Ensure column names are wavelength values
# Example: "400", "405", "410", etc.
```

### Missing Values

**Problem**: `Error: missing values in data`

**Solution**:
```{r fix_missing_values, eval = FALSE}
# Check for missing values
sum(is.na(your_data))

# Remove rows with missing values
your_data <- your_data[complete.cases(your_data), ]

# Or impute missing values
# For spectra, consider interpolation
library(zoo)
your_data[spectral_cols] <- lapply(your_data[spectral_cols], 
                                   function(x) na.approx(x, na.rm = FALSE))
```

## Model Training Issues

### Convergence Problems

**Problem**: Models not converging or poor performance

**Solution**:
```{r fix_convergence, eval = FALSE}
# 1. Check data quality
summary(your_data$target_variable)
hist(your_data$target_variable)

# 2. Try different preprocessing
pp <- st_preprocess(
  data = your_data,
  target = "target_variable",
  preprocess = c("RAW", "ABS", "SNV"),  # Start simple
  noise_range = NULL
)

# 3. Reduce model complexity
fit <- st_run(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS"),  # Single preprocessing
  algos = c("PLSR", "SVR"),  # Fewer algorithms
  outer_k = 3,  # Smaller CV
  inner_k = 3,
  outer_rep = 1,  # Single repetition
  seed = 42
)
```

### Memory Issues

**Problem**: `Error: cannot allocate vector of size X Mb`

**Solution**:
```{r fix_memory, eval = FALSE}
# 1. Reduce dataset size
your_data_small <- your_data[sample(nrow(your_data), 100), ]

# 2. Use fewer preprocessing methods
preprocess_simple <- c("ABS", "SG1")

# 3. Use fewer algorithms
algos_simple <- c("PLSR", "SVR")

# 4. Reduce CV parameters
fit <- st_run(
  data = your_data_small,
  target = "target_variable",
  preprocess = preprocess_simple,
  algos = algos_simple,
  outer_k = 3,
  inner_k = 3,
  outer_rep = 1,
  seed = 42
)
```

### Slow Performance

**Problem**: Models taking too long to train

**Solution**:
```{r fix_performance, eval = FALSE}
# 1. Reduce hyperparameter grids
# Edit R/model.R to use smaller grids:
# PLS_NCOMP_GRID <- c(5, 10, 15)  # Instead of c(5, 10, 15, 20, 25, 30)
# SVR_C_GRID <- c(1, 10)  # Instead of c(0.1, 1, 10, 100)

# 2. Use parallel processing
# Install and load parallel
install.packages("parallel")
library(parallel)

# 3. Reduce CV repetitions
fit <- st_run(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS", "SG1"),
  algos = c("PLSR", "SVR"),
  outer_k = 5,
  inner_k = 5,
  outer_rep = 1,  # Single repetition
  seed = 42
)
```

## Preprocessing Issues

### Savitzky-Golay Parameters

**Problem**: `Error in sgolayfilt: filter length must be odd`

**Solution**:
```{r fix_sg_params, eval = FALSE}
# Ensure window size is odd
preprocess_params <- list(
  SG1 = list(w = 17, p = 2, m = 1),  # w must be odd
  SG2 = list(w = 21, p = 4, m = 2)   # w must be odd
)

# Check parameter validity
w <- 17  # Window size
p <- 2   # Polynomial order
m <- 1   # Derivative order

# w must be odd and > p
if (w %% 2 == 0) stop("Window size must be odd")
if (w <= p) stop("Window size must be greater than polynomial order")
```

### Wavelength Range Issues

**Problem**: `Error: wavelength range not found`

**Solution**:
```{r fix_wavelength_ranges, eval = FALSE}
# Check available wavelengths
wl_cols <- grep("^[0-9]+$", names(your_data), value = TRUE)
wl_values <- as.numeric(wl_cols)
range(wl_values)

# Use valid ranges
pp <- st_preprocess(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS", "SG1"),
  wl_keep = c(400, 2000),  # Within available range
  noise_range = c(700, 705)  # Within available range
)
```

## Visualization Issues

### Plot Not Displaying

**Problem**: `st_plot_best()` not showing plot

**Solution**:
```{r fix_plotting, eval = FALSE}
# 1. Check if fit object is valid
str(fit)

# 2. Ensure plot is printed
p <- st_plot_best(fit)
print(p)  # Explicitly print

# 3. Check for errors in fit
if (is.null(fit$summary)) {
  stop("No results to plot. Check if model training completed successfully.")
}

# 4. Try different plot options
p <- st_plot_best(fit, return_data = TRUE)
head(p)
```

### Empty Results

**Problem**: `fit$summary` is empty or NULL

**Solution**:
```{r fix_empty_results, eval = FALSE}
# 1. Check if training completed
if (is.null(fit$summary)) {
  cat("Model training failed. Check error messages above.\n")
}

# 2. Verify data format
str(your_data)

# 3. Check target variable
summary(your_data$target_variable)

# 4. Try with minimal parameters
fit_minimal <- st_run(
  data = your_data,
  target = "target_variable",
  preprocess = "RAW",
  algos = "PLSR",
  outer_k = 3,
  inner_k = 3,
  outer_rep = 1,
  seed = 42,
  verbose = TRUE
)
```

## Performance Issues

### Poor Model Performance

**Problem**: Low R² or accuracy values

**Solution**:
```{r fix_poor_performance, eval = FALSE}
# 1. Check data quality
# Look for outliers
boxplot(your_data$target_variable)

# 2. Try different preprocessing
preprocess_options <- c("RAW", "ABS", "SNV", "SG1", "MSC")
for (prep in preprocess_options) {
  fit <- st_run(
    data = your_data,
    target = "target_variable",
    preprocess = prep,
    algos = "PLSR",
    outer_k = 5,
    inner_k = 5,
    seed = 42,
    verbose = FALSE
  )
  cat(prep, ":", fit$best$RSQval, "\n")
}

# 3. Check for overfitting
# Compare calibration vs validation metrics
fit$summary[, c("RSQcal", "RSQval")]

# 4. Try different algorithms
algos_options <- c("PLSR", "SVR", "RF", "GLMNET")
for (algo in algos_options) {
  fit <- st_run(
    data = your_data,
    target = "target_variable",
    preprocess = "ABS",
    algos = algo,
    outer_k = 5,
    inner_k = 5,
    seed = 42,
    verbose = FALSE
  )
  cat(algo, ":", fit$best$RSQval, "\n")
}
```

### Unstable Results

**Problem**: High variance in CV results

**Solution**:
```{r fix_unstable_results, eval = FALSE}
# 1. Increase CV repetitions
fit <- st_run(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS", "SG1"),
  algos = c("PLSR", "SVR"),
  outer_k = 5,
  inner_k = 5,
  outer_rep = 5,  # More repetitions
  seed = 42
)

# 2. Use stratified sampling
fit <- st_run(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS", "SG1"),
  algos = c("PLSR", "SVR"),
  outer_k = 5,
  inner_k = 5,
  outer_rep = 3,
  dependent_sampling = "stratified_bin",
  dependent_bins = 5,
  seed = 42
)

# 3. Check for data leakage
# Ensure no target information in spectral data
cor(your_data$target_variable, your_data[, grep("^[0-9]+$", names(your_data))])
```

## Debugging Tips

### Enable Verbose Output

```{r debugging, eval = FALSE}
# Always use verbose = TRUE for debugging
fit <- st_run(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS", "SG1"),
  algos = c("PLSR", "SVR"),
  outer_k = 5,
  inner_k = 5,
  outer_rep = 2,
  seed = 42,
  verbose = TRUE  # Enable progress messages
)
```

### Check Intermediate Results

```{r intermediate_results, eval = FALSE}
# 1. Test preprocessing separately
pp <- st_preprocess(
  data = your_data,
  target = "target_variable",
  preprocess = c("ABS", "SG1"),
  verbose = TRUE
)

# Check preprocessing results
str(pp)
dim(pp$X$ABS)
summary(pp$y)

# 2. Test modeling separately
fit <- st_model(
  X_list = pp$X,
  y = pp$y,
  algos = "PLSR",
  outer_k = 3,
  inner_k = 3,
  outer_rep = 1,
  seed = 42,
  verbose = TRUE
)
```

### Validate Data

```{r validate_data, eval = FALSE}
# Create data validation function
validate_spectral_data <- function(data, target) {
  # Check target exists
  if (!target %in% names(data)) {
    stop("Target column not found")
  }
  
  # Check spectral columns
  spectral_cols <- grep("^[0-9]+$", names(data), value = TRUE)
  if (length(spectral_cols) == 0) {
    stop("No spectral columns found")
  }
  
  # Check for missing values
  if (any(is.na(data[target]))) {
    stop("Missing values in target variable")
  }
  
  # Check for infinite values
  if (any(is.infinite(data[target]))) {
    stop("Infinite values in target variable")
  }
  
  cat("Data validation passed!\n")
  cat("Samples:", nrow(data), "\n")
  cat("Wavelengths:", length(spectral_cols), "\n")
  cat("Target range:", range(data[target]), "\n")
}

# Use validation
validate_spectral_data(your_data, "target_variable")
```

## Getting Help

### Error Messages

When encountering errors:

1. **Read the full error message** - it often contains clues
2. **Check the data format** - ensure it matches requirements
3. **Try minimal examples** - isolate the problem
4. **Use verbose output** - see where the process fails
5. **Check package documentation** - `?function_name`

### Common Error Patterns

```{r error_patterns, eval = FALSE}
# Pattern 1: Column not found
# Error: TARGET_COL %in% names(raw_df) is not TRUE
# Solution: Check column names and target specification

# Pattern 2: Memory issues
# Error: cannot allocate vector of size X Mb
# Solution: Reduce dataset size or model complexity

# Pattern 3: Convergence issues
# Warning: model did not converge
# Solution: Check data quality and preprocessing

# Pattern 4: Package loading
# Error in loadNamespace(x) : there is no package called 'X'
# Solution: Install missing dependencies
```

### Support Resources

1. **Package Documentation**: `help(package = "spectrotune")`
2. **Function Help**: `?st_run`, `?st_preprocess`, etc.
3. **Vignettes**: `browseVignettes("spectrotune")`
4. **GitHub Issues**: Report bugs and request features
5. **Email Support**: Johnbechaalany@gmail.com

## Summary

This troubleshooting guide covers:

1. **Installation Issues**: Package dependencies and permissions
2. **Data Format Problems**: Column names, spectral format, missing values
3. **Model Training Issues**: Convergence, memory, performance
4. **Preprocessing Problems**: Parameter validation, wavelength ranges
5. **Visualization Issues**: Plotting and result display
6. **Performance Problems**: Poor results and instability
7. **Debugging Tips**: Verbose output and validation
8. **Getting Help**: Error patterns and support resources

Remember to:
- Always validate your data before analysis
- Start with simple examples and build complexity
- Use verbose output for debugging
- Check package documentation for function requirements
- Report issues with reproducible examples

## Next Steps

- Explore the [Getting Started](getting-started.html) vignette
- Check out [Advanced Features](advanced-features.html) for more options
- See [Model Comparison](model-comparison.html) for analysis techniques
